---
title: "Capstone Machine Learning Report"
author: "J Slaga"
date: "5/31/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Predicting Airline Flight Delays
Once the data was cleaned up and the initial exploration was complete, an in-depth data analysis was performed using machine learning. Again, our stated goal is to help a traveler optimize their time by avoiding potential delays in travel. The main question I wished to answer was: Could airline flight delays be predicted using given data? Using a dataset from 2015, we could examine how and which of the different factors corresponded to the delayed flights - a supervised problem. The goal then is to find a linear regression model that could predict flight delays.

In the exploration of the data, I found that many airlines made up for shorter departure delays in the air, so for the modeling, ARRIVAL_DELAY was chosen as the dependent variable. After investigating the many variables provided in the dataset, DEPARTURE_DELAY, DAYPART, DAY_OF_WEEK, MONTH, and AIRLINE were chosen as the predictors to examine in the creation of a model. It was observed that examining DEPARTURE_DELAY in conjunction with DAYPART produced a higher Adjusted R-squared value, the main metric I was using to determine the prefered model. 

The following models were used:

```{r  warning=FALSE, message=FALSE, eval=FALSE}
## Model 1
model1 <- lm(ARRIVAL_DELAY ~ DEPARTURE_DELAY + DAYPART, data=late_flights)
summary(model1)
 
## Model 2 ***
model2 <- lm(ARRIVAL_DELAY ~ DEPARTURE_DELAY + DEPARTURE_DELAY:DAYPART, data=late_flights)
summary(model2)

## Model 3
model3 <- lm(ARRIVAL_DELAY ~ DEPARTURE_DELAY + DEPARTURE_DELAY:DAYPART + DAY_OF_WEEK, data=late_flights)
summary(model3)

## Model 4
model4 <- lm(ARRIVAL_DELAY ~ DEPARTURE_DELAY + DEPARTURE_DELAY:DAYPART + DAY_OF_WEEK + MONTH , data=late_flights)
summary(model4)

## Model 5
model5 <- lm(ARRIVAL_DELAY ~ DEPARTURE_DELAY + DEPARTURE_DELAY:DAYPART + DAY_OF_WEEK + AIRLINE, data=late_flights)
summary(model5)
```


Model 5 was chosen as the best fitting model with an Adjusted R-squared value of 0.927. It was tested on data split into training and test sets at 80/20.